---
title: "DNAmFrailty R Notebook"
output: html_notebook
---

## Name: Matt McElheron  

## Email: mcelherm@tcd.ie  

## Date: 05/07/2023  

## Dataset:

> This analysis contains EPIC DNA methylation data matrix 
(Rows:CpGs x Columns:Individuals) and an associated clinical data 
matrix (Rows:Individuals x Columns:Features) including a Rockwood Frailty 
Index, here derived from 31 multi-system age-related  health deficits. 

## Aim:

> introduce a basic pipeine of analysis for TILDA Methylation data, including:
  
  * Data Loading
  * Data QC
  * Epigenetic Clock Derivation (DNAmFrailty)
  

## Packages/Set up

```{r, results='hide', echo=TRUE, warning=FALSE, message=FALSE}
#   Data parsing packages
library(dplyr)
#library(doParallel)
#library(parallel)

#   Bioinformatic packages
library(minfi)
#library(limma)
#library(IlluminaHumanMethylationEPICanno.ilm10b5.hg38)
library(glmnet)
library(caret)

```


## Data Loading/Quality Control Filtering

```{r}

# Path containing IDATs (either a directory of IDAT pairs or a SampleSheet.csv)
idat_dir <- "~/EPIC/IDATs"

# Optional: read SampleSheet if present to carry pheno data
ss_file <- file.path(idat_dir, "SampleSheet.csv")
if (file.exists(ss_file)) {
  targets <- read.metharray.sheet(idat_dir, pattern = "SampleSheet.csv")
  rgSet   <- read.metharray.exp(targets = targets, extended = TRUE, force = TRUE)
} else {
  rgSet   <- read.metharray.exp(base = idat_dir, extended = TRUE, force = TRUE)
}

## ---- Detection p-values & basic QC ----
detP <- detectionP(rgSet)                    # probes x samples

# 1) Drop poor-quality samples: >1% failed probes at p > 0.01
sample_fail_frac <- colMeans(detP > 0.01)
keep_samples <- sample_fail_frac <= 0.01
message(sprintf("Dropping %d/%d samples failing detection p>0.01 threshold.",
                sum(!keep_samples), length(keep_samples)))
rgSet <- rgSet[, keep_samples]
detP  <- detP[, keep_samples, drop = FALSE]

# 2) Drop poor-quality probes: failed in >1% of retained samples at p > 0.01
probe_fail_frac <- rowMeans(detP > 0.01)
keep_probes_detp <- probe_fail_frac <= 0.01
message(sprintf("Dropping %d/%d probes failing detection p>0.01 threshold.",
                sum(!keep_probes_detp), length(keep_probes_detp)))
rgSet <- rgSet[keep_probes_detp, ]

## ---- Remove SNP probes & CpH probes by name (Illumina control content) ----
# Illumina array includes:
# - SNP genotyping probes: names start with "rs"
# - Non-CpG (CpH) targeting probes: names start with "ch"
probe_names <- featureNames(rgSet)
is_rs <- grepl("^rs", probe_names)   # SNP probes -> drop
is_ch <- grepl("^ch", probe_names)   # CpH probes -> drop
drop_by_name <- is_rs | is_ch
if (any(drop_by_name)) {
  message(sprintf("Dropping %d 'rs*' (SNP) and %d 'ch*' (CpH) probes.",
                  sum(is_rs), sum(is_ch)))
  rgSet <- rgSet[!drop_by_name, ]
}

## ---- Quantile normalisation ----
# preprocessQuantile returns a GenomicRatioSet (β & M available)
grSet <- preprocessQuantile(rgSet)

## ---- Remove loci with known SNPs (MAF = 0) at CpG/SBE ----
# This removes any probe with *any* known SNP at the CpG or single-base extension.
# (MAF=0 is most stringent; for practical EWAS, MAF=0.05 is common.)
grSet <- dropLociWithSnps(grSet, snps = c("CpG", "SBE"), maf = 0)

## ---- Remove sex chromosomes ----
anno <- getAnnotation(grSet)
autosomes <- seqnames(anno) %in% paste0("chr", 1:22)
message(sprintf("Dropping %d sex-chromosome probes.", sum(!autosomes)))
grSet <- grSet[autosomes, ]


## ---- Final objects ----
Bvalues <- getBeta(grSet)        # β-values (probes x samples)
sheet <- pData(grSet)            # sample metadata
#ann850k <- getAnnotation(IlluminaHumanMethylationEPICanno.ilm10b5.hg38) %>% 
#  as.data.frame() %>% mutate(cpg = rownames(.))


```


## Elastic Net
> Further hypertuning can be performed for additional parameters, such as 
training/test split, but this will depend on your cohort. 


```{r}

# Parameters
outcome      <- "FI31index_w1_w1"
outcome_name <- "Frailty"
folds        <- 10
split_prop   <- 0.65
model_fam    <- c("gaussian","binomial","poisson")[3]  # pick 1, 2, 3

# --- Prepare data & split ---
Y <- sheet[[outcome]]
set.seed(123)
train_idx <- createDataPartition(Y, p = split_prop, list = FALSE)

X_train <- t(Bvalues[,  train_idx, drop = FALSE])
X_test  <- t(Bvalues[, -train_idx, drop = FALSE])
Y_train <- Y[train_idx]
Y_test  <- Y[-train_idx]

# --- Alpha search (elastic net grid) ---
alpha_grid <- c(1, 0.98, 0.75, 0.50, 0.25, 0.00)

cv_list <- lapply(alpha_grid, function(a) {
  cv.glmnet(x = as.matrix(X_train),
            y = Y_train,
            alpha = a,
            family = model_fam,
            nfolds = folds)
})

# Pick alpha with lowest CV error at lambda.min
cv_cvm_min <- sapply(cv_list, function(cv) min(cv$cvm, na.rm = TRUE))
best_alpha <- alpha_grid[which.min(cv_cvm_min)]
best_cv    <- cv_list[[which.min(cv_cvm_min)]]

# --- Predict (test & full) ---
pred_type <- if (model_fam == "binomial") "class" else "response"

pred_test <- predict(best_cv, newx = as.matrix(X_test), s = "lambda.min", type = pred_type)
pred_full <- predict(best_cv, newx = as.matrix(t(Bvalues)), s = "lambda.min", type = pred_type)

# --- Attach to sheet & mark train/test ---
sheet[[paste0("DNAm", outcome_name)]] <- as.vector(pred_full)

sheet$training_test <- factor(
  ifelse(seq_len(nrow(sheet)) %in% train_idx, "Train", "Test"),
  levels = c("Train","Test")
)

write.csv(sheet, file.path(idat_dir, paste0("SampleSheet_DNAm", outcome_name, ".csv")))

```





